---
title: "Practical Machine Learning Project"
author: "Enric Valencia"
date: "October 19, 2014"
output: html_document
---

Summary
-------
The purpose of this project is to apply Practical Machine Learning to the data provided by fitness monitors in order to predict the type of exercise that the individuals under measurement are doing at the data record time. To do so, a model is fit to the training dataset, and this model is then applied to the test dataset.

Loading and cleaning data
-------------------------
To start with, data is read from the CSV files where it is stored:

```{r echo=TRUE,cache=TRUE}
trainData <- read.csv("pml-training.csv", row.names = 1)
testData <- read.csv("pml-testing.csv", row.names = 1) 
```

After visual inspection of the data, it is observed that it has a lot of NA measurements. Moreover, since there are more than 150 predictors, variance is used as a filter to remove predictors that have nearly zero variance. This way, computation time is reduced by using a lower number of predictors.

```{r echo=TRUE,cache=TRUE}
library(caret)

zeroVars <- nearZeroVar(trainData, saveMetrics = T)
trainData <- trainData[, !zeroVars$nzv]
```

Now that the number of predictors has been reduced, modelling can begin.

Training the model
------------------
The kind of problem faced here, is a good candidate for a prediction model based on classification. From all of them, Random Forest is generally very accurate, so this is the first option to be explored here. Furthermore, to avoid overfitting, 10-fold cross-validation is used:

```{r echo=TRUE,cache=TRUE}
randomForestFit <- train(classe ~ ., data = trainData, method = "rf", trControl = trainControl(method = "cv", number = 10))
```
The training process results in the following final model:
```{r echo=TRUE,cache=TRUE}
randomForestFit
randomForestFit$finalModel
```

- Model accuracy: 1
- Out-of-sample error: 0.04%

As it is reported by the model, its 1 accuracy is very good, and its 0,04% out-of-sample error is very low. With these results, the resulting model can be confidently used to predict the exercise type from the test data.
 
Predicting on the test data
---------------------------
Now, the obtained Random Forest model is  used for prediction on the test data:

```{r echo=TRUE,cache=TRUE}
predictedClasses <- predict(randomForestFit,testData)
predictedClasses
```

Finally, and for assignment submission, the prediction output is converted into a characters' array, and the required files are generated:

```{r echo=TRUE}
answers <- as.character(predictedClasses)

n = length(answers)
for(i in 1:n){
  filename = paste0("./subfiles/problem_id_",i,".txt")
  write.table(answers[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
```